## import packages
```{r}
library(tidyverse)
library(reshape2)
library(countrycode) #package to convert country codes
```

## set the working directory and read file
```{r}
knitr::opts_chunk$set(root.dir = normalizePath("C:/Users/David Jhang/Desktop/NYC Data Academy/Project 2/Data/"))
df = read.csv("Data/master.csv",stringsAsFactors = FALSE)
```

## check for NAs
- What is HDI? (seems like a lot of missing data, removing for that reason)
```{r}
colSums(is.na(df))
sum(is.na(df$HDI.for.year)) #19456 NAs.. remove
df$HDI.for.year = NULL

```

## checking column names and renaming for formatting issues
```{r}
df = df %>%
  rename(country = `ï..country`,
         num_suicides = suicides_no,
         suicides_100k = suicides.100k.pop,
         key = country.year,
         gdp = `gdp_for_year....`,
         gdp_per_capita = `gdp_per_capita....`)

glimpse(df)

```
## Check data integrity
- The data is shown with an age group (6 groups) and split genders (2) = 6*2
- There should be 12 rows for each year
- want to make sure there is any skewed data (missing gender or age groups)
- remove anything that doesn't total 12
```{r}
df %>%
  group_by(key) %>%
  count() %>%
  filter(n != 12)

#looks like it's only 2016 that has this issue, remove
df = df %>%
  filter(year != 2016)

```

## Check the date range is okay to use
- I noticed there was an issue that only a small # of countries provided data for the entire 1985-2015 period.
- Wanted to include all major countries and make sure a good # of countries were provided for the entire analysis
- I decided to choose the year range of 2002-2015 which included 52 countries out of the 101
```{r}
#check all the country values
unique(df$country)

#this will only keep data with year from 2000-2015
df = df %>%
  filter( between(year,2002,2015) )
unique(df$year) #checks to make sure the filter worked correctly

# 16 years * 12 rows for each year = 192 .. created variable n for this
df = df %>%
  group_by(country) %>%
  mutate(count=n())

# delete any countries that does not have 192
df = df %>% 
  filter( count == 168)

# delete the count column
df$count = NULL

#check to see data is okay
df %>%
  group_by(country) %>%
  summarize(n = n(), years = n / 12) %>%
  arrange(years)

```

## Exploring the Dataset
  - Need to clean "years" from $age and also white spaces
  - Need to remove commas and turn gdp into an num
```{r}

df = df %>% 
  mutate(age = str_remove(age,"years")) #remove yeaqrs
df = df %>% 
  mutate(age = str_remove(age," ")) #remove whitespaces

df$gdp <- as.numeric(gsub(",","",df$gdp)) #remove and convert gdp to numeric

```

## Found a way to enter Region Code into my dataset
- could be useful to group by region later
```{r}

?codelist # needed to match the origin and the destination

#takes country column and uses that to reference and find a continent
df = df %>% 
  mutate(region = countrycode(sourcevar = country, 
                              origin = "country.name",
                              destination = "region"))

glimpse(df) #check the types
unique(df$country) #52 countries
unique(df$year) #2002-2015 year range
unique(df$region) #6 regions now

```

## Writing cleaned dataframe into .csv format
```{r}
write.csv(df, "df.csv", row.names=FALSE)

```

