## import packages
```{r}
library(tidyverse)
library(reshape2)
library(ggplot2)
library(hrbrthemes)
library(ggalt)
library(ggpubr)
library(gridExtra)
library(grid)
```

## set the working directory and read file
```{r}
knitr::opts_chunk$set(root.dir = normalizePath("C:/Users/David Jhang/Desktop/NYC Data Academy/Project 2/Data/"))
df = read.csv("df.csv",stringsAsFactors = FALSE)
```

## Graph 0
```{r}

#Globally suicide rate over the years
# decreasing trend
graph0 = df %>%
  group_by(year) %>%
  summarise(year_suicides = sum(num_suicides), year_pop = sum(population), ratio = (year_suicides/year_pop) *100000 )

graph0

graph0a = df %>%
  group_by(year, sex) %>%
  summarise(year_suicides = sum(num_suicides), year_pop = sum(population), ratio = (year_suicides/year_pop) *100000 )

ggplot(graph0, aes(x = year, y = ratio, color= ratio)) +
  geom_point() +
  geom_line() +
  labs(title = "Global suicide rates over the years", x = "Year", y = "Suicides (100k)")

# conclusion = suicide rates are going down globally, most likely not for each country

```
## Graph0A
- standard deviation graph of each country from 2002-2005
```{r}

# creating new df for each country of total suicides/100k that year
test = df %>%
  group_by(year, country) %>%
  summarise(total_suicides = sum(num_suicides), total_pop =sum(population))
test = test %>%
  group_by(year, country) %>%
  summarise(suicides100k = (total_suicides/total_pop) * 100000)

test2002 = test %>%
  filter(year == 2002)
test2003 = test %>%
  filter(year == 2002)
test2004 = test %>%
  filter(year == 2002)
test2005 = test %>%
  filter(year == 2002)
test2006 = test %>%
  filter(year == 2002)
test2007 = test %>%
  filter(year == 2002)
test2008 = test %>%
  filter(year == 2002)
test2009 = test %>%
  filter(year == 2002)
test2010 = test %>%
  filter(year == 2002)
test2011 = test %>%
  filter(year == 2002)
test2012 = test %>%
  filter(year == 2002)
test2013 = test %>%
  filter(year == 2002)
test2014 = test %>%
  filter(year == 2002)
test2015 = test %>%
  filter(year == 2015)

test1= test2002 %>%
  left_join(test2003, by = "country") %>%
  rename("suicides2002" = "suicides100k.x", "suicides2003" = "suicides100k.y")
test1$year.x = NULL
test1$year.y = NULL


#### NEED TO RUN THIS FOR EACH YEAR ##### test2003, test2004, etc..
# iterate this by each year to pull each suicide100k into a separate column
# change test2015 and suicides2015 respectively
test1= test1 %>%
  left_join(test2015, by = "country") %>%
  rename("suicides2015" = "suicides100k")
test1$year=NULL

#creating a new df with all the columns
test_final = test1

#run SD by row and create new column to graph
test_final = transform(test_final, SD=apply(test_final,1, sd, na.rm = TRUE))
test_final

ggplot(test_final, aes(x=reorder(country, SD), y=SD)) +
  geom_bar(stat="identity", fill="steelblue") +
  geom_hline(yintercept = mean(test_final$SD), color="red", linetype = "dotted") +
  labs(title = "Standard Deviation of Suicides (100k) by Country", x = NULL, y = "Standard Deviation") +
  coord_flip()

#would expect this because I assume that some years would be good years or bad years would elevate suicides (financial crisis)

```


## Graphs 1
  - countries vs suicide per population (100k)
```{r}

# can't use mean of suicides_100k since this is all weighted by population
graph1 = df %>%
  group_by(country, region) %>%
  summarise(suicide100k = sum(num_suicides) / sum(population) * 100000)
  
graph1

ggplot(graph1, aes(x=reorder(country, suicide100k), y=suicide100k, fill=region)) +
  geom_bar(stat="identity") +
  geom_hline(yintercept = mean(graph1$suicide100k), color="red", linetype = "dotted") +
  scale_fill_manual("Region", values = c("East Asia & Pacific"="#cd7058",
                                         "Europe & Central Asia"="#599ad3",
                                         "Latin America & Caribbean" = "#727272",
                                         "Middle East & North Africa" = "#79c36a",
                                         "North America" = "#9e66ab",
                                         "Sub-Saharan Africa" = "#f9a65a")) +
  labs(title = "Suicides from 2002-2015", x = "Country", y = "Suicides (100k)") +
  coord_flip()

# top/least suicide rates by country

```
## Graphs 2
  - By gender and continent
```{r}

# new df by gender and suicide 100k
graph2 = df %>%
  group_by(sex, region) %>%
  summarise(suicide100k = sum(num_suicides) / sum(population) * 100000)
  
#dodge to show the diference between male vs female
ggplot(graph2, aes(x=region, y=suicide100k, fill=sex)) +
  geom_bar(stat="identity", position="dodge") +
  labs(title = "Male/Female Suicides by Region (per 100k)", x = "Region", y = "Suicides (100k)")

# appears that males commit suicide more than females

```
## Graphs 3
  - wanted to visualize the ratio since from graph 2 it appeared that males have higher risk to suicide
  - Proportion of male vs female suicides per country
```{r}

# creating column to create a ratio between genders
test = df %>%
  group_by(country, sex) %>%
  summarise(split  = sum(num_suicides))

all_suicides = df %>%
  group_by(country) %>%
  summarise(all_suicides=sum(num_suicides))

#checking new columns
test
all_suicides

# join the 2 new columns by country
graph3 = test %>%
  left_join(all_suicides, by = c('country'='country'))

# now trying to find the male ratio so we can group by in the graph
graph3 = graph3 %>%
  group_by(country, sex) %>%
  summarise(ratio = split/all_suicides)

graph3a = graph3 %>%
  pivot_wider(country, names_from=sex, values_from=ratio) %>%
  mutate(gender_ratio = male/female)

#actual graph  
graph3a %>%
  left_join(graph3, by="country") %>%
  ggplot(aes(x=reorder(country, gender_ratio), y=ratio, fill=sex)) +
  geom_bar(stat="identity") +
  labs(title = "Suicides from 2002-2015", x = "Country", y = "Ratio by Gender") +
  geom_hline(yintercept = mean(graph3a$male), color="blue", linetype = "dotted") +
  geom_hline(yintercept = mean(graph3a$female), color="red", linetype = "dotted") +
  coord_flip()

#conclusion = males typically are higher risk to suicides

```
## Graph4a
  - Wanted to show which age group had the highest.  Hunch that the elderly have a higher suicide rate
```{r}

#new df for age group analysis
graph4a = df %>%
  group_by(age) %>%
  summarise(suicide100k = sum(num_suicides) / sum(population) * 100000) %>%
  arrange(desc(suicide100k))

graph4a

#factorizing age to sort when charting
graph4a$age = factor(graph4a$age, levels=c("5-14","15-24","25-34","35-54","55-74","75+"))

ggplot(graph4a, aes(x=age, y=suicide100k, fill=age)) +
  geom_bar(stat="identity") +
  scale_fill_manual("Age", values = c("5-14" = "royalblue2",
                                      "15-24"="paleturquoise1",
                                      "25-34"="lightsteelblue1",
                                      "35-54" = "steelblue1",
                                      "55-74" = "steelblue3",
                                      "75+" = "steelblue4")) +  
  labs(title = "Suicides by Age Group", x = "Age Group", y = "Suicides (100k)")

## conclusion = elderly has the highest suicide rate, sad to see but trends upwards

```


## Graphs 4
  - By age group 
  (wanted to do country but graph might be too big so will do it by continent)
```{r}

#this time wanted to add region to the df
graph4 =df %>%
  group_by(age, region) %>%
  summarise(suicide100k = sum(num_suicides) / sum(population) * 100000) %>%
  arrange(age)


#factorizing the age to help with sorting  when making a chart
graph4$age = factor(graph4$age, levels=c("5-14","15-24","25-34","35-54","55-74","75+"))
graph4

# needed to change colors since it was too confusing to view but wanted to show them with dodge(side-by-side)
ggplot(graph4, aes(x=region, y=suicide100k, fill=age)) +
  geom_bar(stat="identity", position="dodge") +
  scale_fill_manual("Age", values = c("5-14" = "royalblue2",
                                      "15-24"="paleturquoise1",
                                      "25-34"="lightsteelblue1",
                                      "35-54" = "steelblue1",
                                      "55-74" = "steelblue3",
                                      "75+" = "steelblue4")) +  
  labs(title = "Suicides by Age Group", x = "Country", y = "Suicides (100k)")

#conclusion, Asia appears to have an issue with suicide

```

#Graph 5 - Suicide by Year (by Generation)
- no real trends here
```{r}

#creating new df for generation analysis vs suicides per 100k
graph5 = df %>%
  group_by(generation, year) %>%
  summarise(suicide100k = sum(num_suicides) / sum(population) * 100000)

graph5

#doing an overlay to show it all on one chart
ggplot(graph5, aes(year, suicide100k, group=generation, colour=generation, fill=generation)) + 
  geom_point(shape=21, size=3) +
  geom_line() +
  labs(title = "Yearly Suicide Rates by Age Group", x = "Year", y = "Suicides (100k)")


```
#Graph 6
- Diverging Bars graph representing the % change from 2002 vs. 2015
```{r}

# getting data from 2002 and 2015 to see the % change of suicides
test = df %>%
  group_by(year, country) %>%
  summarise(total_suicides = sum(num_suicides), total_pop =sum(population))
test = test %>%
  group_by(year, country) %>%
  summarise(suicides100k = (total_suicides/total_pop) * 100000)

test2002 = test %>%
  filter(year == 2002)
test2015 = test %>%
  filter(year == 2015)

test1= test2002 %>%
  left_join(test2015, by = "country") %>%
  rename("suicides2002" = "suicides100k.x", "suicides2015" = "suicides100k.y")
test1$year.x = NULL
test1$year.y = NULL
test1


graph6 = test1 %>%
  group_by(country, suicides2002, suicides2015) %>%
  summarise(change = ((suicides2015 - suicides2002) / suicides2002) * 100) %>%
  arrange(change)

## Grenada had 0 in 2002 and 2015 so that's why it's empty.  Population of 112k

graph6


ggplot(graph6, aes(reorder(x=country,change), y=change, col = change)) + 
  geom_point(stat='identity', size = 4) +
  geom_hline(yintercept = 0, col = "grey", size = 1) +
  scale_color_gradient(low = "tomato", high = "royalblue") +
  geom_segment(aes(y = 0, 
                   x = country, 
                   yend = change, 
                   xend = country), size = 1) +
  labs(title="% Change of Suicides", 
       x = NULL, y = "% change in suicides per 100k (2002 vs 2015)") +
  scale_y_continuous(breaks = seq(-80, 80, 10), limits = c(-80, 80)) +
  theme_gray(base_size=14) +
  theme(legend.position = "none") +
  coord_flip()
```

Graph 6a:
- first, trying to make a df with average gdp(capita) and total_suicides
- then run correlation to see which countries have the highest and lowest rates of suicide every year
```{r}

#creating df with average of gdp and # suicides
corr = df %>%
  group_by(country, year) %>%
  summarise(gdp_per_capita = mean(gdp_per_capita), total_suicides=round(sum(num_suicides),2))

#found a way to basically run correlation for every year for both gdp and suicides
# this represents countries that are getting richer over the years and increasing # of suicides

corr1 = corr %>%
  group_by(country) %>%
  summarise(gdp_corr = cor(year, gdp_per_capita), suicides_corr = cor(year, total_suicides)) %>%
  arrange(desc(suicides_corr))

#USE THIS FOR PRESENTATION
#so basically this is saying every year as people were getting wealthier on average.. there was a increase in suicides every year
corr1 

ggplot(corr1, aes(reorder(x=country,suicides_corr), y=suicides_corr, col = gdp_corr)) + 
  geom_point(stat='identity', size = 4) +
  geom_hline(yintercept = 0, col = "grey", size = 1) +
  scale_color_gradient(low = "tomato", high = "royalblue") +
  geom_segment(aes(y = 0, 
                   x = country, 
                   yend = suicides_corr, 
                   xend = country), size = 1) +
  labs(title="Correlation of Total Suicides (per Year)", 
       x = NULL, y = "correlation") +
  scale_y_continuous(breaks = seq(-1, 0, 1), limits = c(-1, 1)) +
  theme_gray(base_size=14) +
  theme(legend.position = "none") +
  coord_flip()

```

#Graph 7: GDP vs Suicides
Does countries with higher GDP have less suicides?
```{r}

#creating new df for GDP vs Suicide analysis
# mean of gdp_per_capita should hold the value of gdp since its the same value for each year
test = df %>%
  group_by(year, country) %>%
  summarise(suicides = sum(num_suicides), population = sum(population), gdp_per_capita=mean(gdp_per_capita))

graph7 = test %>%
  group_by(year,country, gdp_per_capita) %>%
  summarise(suicides100k = (suicides/population) * 100000)

ggplot() +
  geom_point(data = graph7,aes(x = gdp_per_capita, y = suicides100k, colour = country)) +
  labs(title = "Average suicides (100k) vs. GDP (capita) yearly by Country ", x = "GDP (capita)", y = "Suicides (100k)") +
  theme_gray(base_size=14) +
  theme(legend.position = "none") +
  xlim(0,80000)

#too noisy.. yearly may not be the best option here

```
# Graph 7a
- Instead of year let's just do it by region
```{r}
#this time just group by country and region and should give the mean of that period
graph7a = df %>%
  group_by(country, region) %>%
  summarize(suicides100k = sum(num_suicides) / sum(population) * 100000, 
            gdp_per_capita = mean(gdp_per_capita))
graph7a

ggplot() +
  geom_point(data = graph7a,aes(x = gdp_per_capita, y = suicides100k, colour = region)) +
  theme_gray(base_size=14) +
  labs(title = "Average suicides (100k) vs. GDP (capita) from 2002-2015 ", x = "GDP (capita)", y = "Suicides (100k)")

```

## Now we choose the US since it had the highest correlation and then want to do linear regression on 
#Graph 8
- I selected the US, one of the top offenders, clearly meaningful for us to conceptualize
```{r}

# plotting yearly suicide increase over the past year in the US
graph8 = df %>%
  filter(country=="United States") %>%
  group_by(year) %>%
  summarise(yearly_suicides = sum(num_suicides), gdp_cap = mean(gdp_per_capita))
graph8

# linear and positive
ggplot(graph8, aes(x = year, y = yearly_suicides, color= yearly_suicides)) +
  geom_point() +
  geom_line() +
  xlim(2002,2016) +
  labs(title = "Yearly Suicides in the US", x = "Year", y = "# of Suicides") + 
  theme(legend.position = "none")

```

#Graph9
- over the years the total_suicides vs gdp_per_capita in the US
```{r}
corr
# using the corr dataframe that I made that highlights gdp_per_cap and total_suicides every year, will select US and graph
us = corr %>%
  filter(country=="United States") %>%
  group_by(year) 

#will try a regression model on this one later
cor(corr$gdp_per_capita,corr$total_suicides)

ggplot(us, aes(x = total_suicides, y = gdp_per_capita, color= total_suicides)) +
  geom_point() +
  geom_line() +
  labs(title = "Yearly # of Suicides vs Yearly GDP (US)", x = "Yearly Suicides", y = "GDP (per capita)") +
  theme(legend.position = "none")

```
#Graph 10
- M/F split for US
```{r}

# create new DF for just US to try linear regression for US only
df1 = df %>%
  filter(country=="United States") %>%
  group_by(year, sex) %>%
  summarise(yearly_suicides = sum(num_suicides), yearly_pop= sum(population), gdp_cap=mean(gdp_per_capita))

ggplot(df1, aes(x = sex, y = yearly_suicides, fill = sex)) +
    geom_boxplot() +
    geom_jitter(shape = 15,
        color = "steelblue",
        position = position_jitter(0.21)) +
    labs(title = "Male vs Female Breakdown in the US", x = "Gender", y = "# of suicides (yearly)") +
    theme_classic()

```

```{r}
# create new DF for just US to try linear regression for US only
df2 = df %>%
  filter(country=="United States") %>%
  group_by(year, age) %>%
  summarise(yearly_suicides = sum(num_suicides), yearly_pop= sum(population), gdp_cap=mean(gdp_per_capita))

ggplot(df2, aes(x = age, y = yearly_suicides, fill = age)) +
    geom_boxplot() +
    geom_jitter(shape = 15,
        color = "steelblue",
        position = position_jitter(0.21)) +
    labs(title = "Yearly Suicides in the US by Age Group", x = "Age Group", y = "# of suicides (yearly)") +
    theme_classic()

```